Molecular dynamics simulation of synchronization in driven particles

### 2020-11-24 manuscript/manuscript_ajp_2020.tex
updated some typos

% The % character begins a comment, which continues to the end of the line.

\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures
\usepackage{pythontex}

@@

\section{Molecular Dynamics Model}
\label{sec:MD}
We use a classical two-dimensional model for 
studying the dynamics of $N$ interacting particles. 
Particles are confined in a two-dimensional (2D) 
simulation of area $A = L \times L$ where $L=36.5 a_0$
where $a_0$ is a dimensionless unit of length.
An individual particle $i$ has
position $\vec{r}_i = x_i \hat{x} + y_i \hat{y}$.
Particles are subject to
periodic boundary conditions
such that a particle leaving the edges of the system is mapped
back to a position within the simulation 
by $x_i+L=x_i$ and $y_i+L=y_i$.
We summarize the units of the simulated variables in Table ~\ref{tab:1}.

We model particle interaction forces
$\vec{F}_{ij} = -\nabla V_{ij}(r_{ij})$ 
with
the Yukawa potential energy 
\begin{equation}
  V_{ij}(r_{ij}) = \frac{E_0}{r_{ij}} e^{-\kappa r_{ij}},
\end{equation}
where particle $i$ and $j$ are distance
$r_{ij} = |\vec{r}_i - \vec{r}_j|$ apart.
This 
screened Coulomb potential
%$E_0=2$ scales strength of repulsion
is scaled in terms of energy unit $E_0$
defined in Table ~\ref{tab:1}.
%$ = k q_1 q_2$. %[CHECK scaling/units].
$\kappa = 1/R_0$ is the screening parameter 
that describe the lengthscale at
which particles interact.
We fix $R_0$ to be $a_0$.
In experiments charge screening is observed
due to ions in the suspending fluid and
the charges of surrounding particles
which
reduces the interaction range of individual particles.
This allows the numerical models to run more efficiently
since particles do not need to interact over long range.
[explain particle in cell models].

@@

@@

We use several model enviroments to confine the particles,
again assuming the confining force arises from a potential function 
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 $\vec{F}_{l}(\vec{r}) = -\nabla V_l(\vec{r}).$
 The landscape potential $V(\vec{r}) $ are static 
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}.

 One example of a periodic landscape is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l(y) = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of periods,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

 In multi-particle simulations, we confine
 the particles along the $x-$direction with a
 quasi-one-dimensional geometry
 using a periodic function.
 \begin{equation}
   V_{q1D}(x) = V_0 \cos{(\pi x / L)}
 \end{equation}.
 %%We could also use a Gaussian function.

The simulation is controlled by a $for()$ loop
which runs from an initial to maximum $time$ integer.
Each integer timestep
represents a simulation time of $\Delta t=0.001$.
At each timestep
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then integrate
the equation of motion to move particles
to an updated position
$\vec{r}_i(t+\Delta t)$.
%
The integration is simple
because 
we model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
Here 
the suspending fluid is highly viscous
and exerts a damping force on the particles equivalent
to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that 
The damping comes from a viscous fluid model
providing a nonconservative force,
modeled as a linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity is high [finish... reference...]
This model is encountered in standard texts,
for instance Ch. 2 of Taylor's Classical Mechanics.

A particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location in the simulation
with respect to other particles $j$.
Since we model the acceleration as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

@@

The hopping pattern of the driven particle is periodic,
and could be achieved over a range of
driving forces.  %, as we will demonstrate in Fig.~\ref{fig:1}
We explore the ranges of periodic hopping patterns
by increasing $F_{DC}$ as a function of time,
as shown in 
%
Fig.~\ref{fig:1}.
%we show the result increasing the 
%force to the particle as a function of time.  
%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of $\Delta F_{DC} = 0.001$ every $\Delta t = 4000$ integer timesteps
or $40$ time units.
%this seems fast, but it is slow enough to estimate
%the velocity accurately.
%this is not be so for a multiple particle system!
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

>>>>

% The % character begins a comment, which continues to the end of the line.
\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures
\usepackage{pythontex}

@@

\section{Molecular Dynamics Model}
\label{sec:MD}
We use a classical two-dimensional model for 
studying the dynamics of $N$ interacting particles. 
Particles are confined in a two-dimensional (2D) 
simulation of area $A = L \times L$ where $L=36.5 a_0$
where $a_0$ is a dimensionless unit of length.
An individual particle $i$ has
position $\vec{r}_i = x_i \hat{x} + y_i \hat{y}$.
Particles are subject to
periodic boundary conditions
such that a particle leaving the edges of the system is mapped
back to a position within the simulation 
by $x_i+L=x_i$ and $y_i+L=y_i$.
The units of the simulated variables are in Table ~\ref{tab:1}.

@@

We model particle interaction forces
$\vec{F}_{ij} = -\nabla V_{ij}(r_{ij})$ 
with
the Yukawa potential energy 
\begin{equation}
  V_{ij}(r_{ij}) = \frac{E_0}{r_{ij}} e^{-\kappa r_{ij}},
\end{equation}
where particle $i$ and $j$ are distance
$r_{ij} = |\vec{r}_i - \vec{r}_j|$ apart.
This 
screened Coulomb potential
%$E_0=2$ scales strength of repulsion
is scaled in terms of energy unit $E_0$
defined in Table ~\ref{tab:1}.
%$ = k q_1 q_2$. %[CHECK scaling/units].
$\kappa = 1/R_0$ is the screening parameter 
that describes the lengthscale at
which particles interact.
We fix $R_0$ to be $a_0$.
In experiments charge screening is observed
due to ions in the suspending fluid and
the charges of surrounding particles
which
reduces the interaction range of individual particles. % \cite{}.
Because the particles interact over short ranges, 
the numerical models can be run efficiency
using a neighbor list algorithm
determined using a cell method.
[explain!]

@@

We use several model enviroments to confine the particles,
assuming the confining force arises from a potential function 
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 $\vec{F}_{l}(\vec{r}) = -\nabla V_l(\vec{r}).$
 The landscape potential $V(\vec{r}) $ are static 
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}.
% 
% One example of a periodic landscape is
%% %that used in Sec.~\ref{sec:one}
% where 
% \begin{equation}
%   V_l(y) = V_0 \cos{(N_p \pi y / L)}
% \end{equation}
% where $N_p$ are the number of periods,
% and $V_0$ is an adjustable parameter
% to set the depth of the troughs.
%
 In multi-particle simulations, we confine
 the particles along the $x-$direction 
 using a periodic function 
 \begin{equation}
   V_{q1D}(x) = V_0 \cos{(\pi x / L)}
 \end{equation}.
 %%We could also use a Gaussian function.
 This channel potential keeps the particles
 interacting at short range confined primarily to motion
 along the y-direction
 but allows for some lateral motion of particles.
 This is sometimes referred to as a
 quasi-one-dimensional geometry.

The simulation is controlled by a $for()$ loop
which runs from an initial to maximum $time$ integer.
Each integer timestep
represents a simulation time of $\Delta t=0.001$.
At each timestep
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then integrate
the equation of motion to move particles
to an updated position
$\vec{r}_i(t+\Delta t)$.
%
The integration is simple
because 
we model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
%The suspending fluid is highly viscous
%and exerts a damping force on the particles equivalent
%to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that 
The damping comes from a viscous fluid model
providing a nonconservative force,
modeled as a linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity is high [finish... reference...]
This model should be familiar to readers
modeling Milliken's oil drops
in a standard classical mechanics text \cite{Taylor2005}.
%This model is encountered in standard texts,
%for instance Ch. 2 of Taylor's Classical Mechanics.

A particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location in the simulation
with respect to other particles $j$.
Since the acceleration as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

@@

The hopping pattern of the driven particle is periodic,
and could be achieved over a range of $F_{DC}$.  %, as we will demonstrate in Fig.~\ref{fig:1}
We explore the ranges of periodic hopping patterns
by increasing $F_{DC}$ as a function of time,
as shown in 
%
Fig.~\ref{fig:1}.
%we show the result increasing the 
%force to the particle as a function of time.  
%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of $\Delta F_{DC} = 0.001$ every $\Delta t = 4000$ integer timesteps
or $40$ time units.
%this seems fast, but it is slow enough to estimate
%the velocity accurately.
%this is not be so for a multiple particle system!
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.
